Coordinated monitoring of the Piney Point wastewater discharge into Tampa Bay: Data synthesis and reporting

### 2022-05-24 manu-draft.Rmd
proofing full draft

The long-term recovery of Tampa Bay, Florida represents a collaborative, multi-stakeholder effort to quantify and limit external nitrogen loading as a key stressor impacting ecosystem health.  Nitrogen has been reduced by approximately 2/3 of historical loading primarily through upgrades to wastewater treatment systems, stormwater controls, and habitat restoration projects [@Johansson91;@Lewis98;@Greening14;@Beck19].  As a result, chlorophyll concentrations have decreased, water clarity has improved, and seagrasses recovered to a historical 1950s baseline by 2016 [@Greening14;@Sherwood17].  Despite these historical gains, ongoing management challenges for protecting and restoring the natural resources of Tampa Bay persist from continued threats of climate change and increasing watershed development.  Recent seagrass losses and seasonal harmful algal blooms highlight the potential impacts of persistent and increasing stressors affecting the bay [@Lopez16;@Tomasko20].

Impacts to human health and the environment associated with extractive mining activities are pervasive worldwide.  In Florida, phosphate mining has generated a multi-billion dollar fertilizer industry that supports agricultural production on the global market [@Henderson04]. Seventeen phosphate mining facilities occur in the Tampa Bay watershed and wastewater release into surface waters from these facilities has negatively impacted environmental resources.  The Piney Point facility on the southeast shore of Tampa Bay has been inactive for over twenty years and holding capacity to maintain wastewater onsite has decreased over time, contributing to releases in the early 2000s and in 2011 to Tampa Bay [@Garrett11;@Switzer11].  In March 2021, a leak was identified in the southern holding pond and wastewater release was authorized by the Florida Department of Environmental Protection to prevent catastrophic failure of the holding ponds.  Approximately 215 million gallons of mixed legacy phosphate mining wastewater and seawater from port dredging operations were discharged over a ten day period, introducing 205 tons of nitrogen to Tampa Bay [@Beck22].

@@

# Data collection and synthesis

Monitoring of bay resources in response to release from Piney Point began in late March of 2021 and continued through the summer and early fall.  Several partners from public, private, and academic institutions conducted monitoring of water quality, phytoplankton (including the red tide organism *Karenia brevis*), seagrasses, macroalgae, and benthic habitats (Figure \@ref(fig:map)).  Monitoring partners included the Florida Department of Environmental Protection (FDEP), Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute (FWRI) of the Florida Fish and Wildlife Conservation Commission (FWC), City of St. Petersburg, Tampa Bay Estuary Program (TBEP), Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida (Figure \@ref(fig:map)a).  Coordination among monitoring agencies was facilitated by the TBEP under guidance of a plume simulation model run by the Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science [@Chen18;@Chen19]. Initial monitoring was focused near the release site at Port Manatee (Figure \@ref(fig:map)), but later expanded throughout middle and lower Tampa Bay, Boca Ciega Bay, Manatee River, and Sarasota Bay, as the simulation model predicted dispersal of the plume away from Piney Point [@Liu21].  Sampling effort was most intense in April, with less frequent sampling from May through September, although seagrass and macroalgae sampling was consistent throughout the event (Table \@ref(tab:efftab)).

Water quality monitoring focused on parameters that were expected to indicate a response to elevated nutrient loads from the Piney Point release.  Because Tampa Bay is a nitrogen-limited estuary [@Greening14;@Greening06;@Wang99], water quality monitoring focused primarily total nitrogen and inorganic forms (i.e., total ammonia nitrogen, nitrate/nitrite), chlorophyll-a as an indication of phytoplankton growth, and additional parameters expected to respond to nutrient loading (e.g., secchi depth, dissolved oxygen).  Phytoplankton monitoring included both qualitative (presence/absence) and quantitative (cells/L) summaries of taxa enumerated by microscopy from water samples taken at similar frequency and distribution as the water quality sites.  Phytoplankton data also included event-based sampling for *Karenia brevis*.  Finally, seagrass and macroalgae data were collected by transect sampling at locations covering the shoreline north of Piney Point and extending to the south into northern Sarasota Bay (Figure \@ref(fig:map)b). Transect sampling included identification of seagrass species and macroalgae, typically to genus, and Braun-Blanquet abundance estimates at 10 m distances along each 50 m transect.  Details on all sampling methods are provided in @Beck22.

@@

Combining the source files on Google Drive into a synthesized format in R also required a consistent set of variables and naming convention that were applied to each file.  For example, parameters included in each water quality file were converted to a common set of names and all units were standardized (e.g., total nitrogen as `tn` and mg/L, chlorophyll-a as `chla` and $\mu$g/L, etc.). This required developing custom processing code for data files from each partner to address each unique use case.  As a result, the final processing file included almost 3000 lines of code to create the synthesized data products.  Each partner agency was also assigned a unique identifier in the synthesized files so effort could be tracked and reported in the dashboard (e.g., Pinellas County as `pinco`, Florida Department of Environmental Protection as `fldep`, etc.).  All synthesized files were saved in a long and tidy format, where each observation was given its own row, each variable was in its own column, and only one value was assigned to a cell [@Wickham14c].  This allowed for easily filtering observations of interest (as required with the dashboard) using the standardized naming conventions applied to each source file. More importantly, the developed workflow was used iteratively, often several times each week as new data files were uploaded to the Drive folders.  This required executing the synthesis code locally before pushing data to the dashboard and constantly revising the code to address new data formats or issues encountered when new files were uploaded.

The local workflow created by TBEP was hosted online in a GitHub repository [https://github.com/tbep-tech/piney-point, @Beck21d] that allowed use of additional resources that were critical to the synthesis effort (Figure \@ref(fig:synthesis), testing and sharing).  Hosting the RStudio project on GitHub served multiple purposes of 1) using version control to document and track changes using Git, 2) making the code publicly accessible for transparency with the larger research and management community, 3) allowing automated testing of repository content, and 4) serving as an access point for deploying the dashboard on Shiny Server (see below).  The use of automated testing was an important service to identify potential errors or mistakes in the synthesized datasets. The testthat R package [@Wickham11] was used to write several checks for each dataset that were considered essential for maintaining integrity of the information.  Each time changes were made to the workflow and pushed to the GitHub repository, tests were run automatically with GitHub Actions to verify each dataset included appropriate and accurate results.  For example, a test was used to verify that all water quality parameters were appropriately named and did not include any missing values. A notification was sent via email to TBEP staff if an error was encountered.  This automated service reduced the amount of time required to manually check the data and provided assurance that the reported results in the dashboard were correct.

@@

An important service provided by the dashboard was creating context for the response-based monitoring results relative to baseline conditions in Tampa Bay.  This allowed dashboard users to quickly assess if the 2021 conditions were abnormal relative to typical seasonal patterns observed in the bay.  In addition to the baseline data tabs, all water quality data from the response-based monitoring were compared to the long-term monthly averages for data at the nearest ambient monitoring station [@Beck22].  Elements of the UI and summary plots allowed a user to quickly determine if the results were outside of the expected range.  Map points for a sampling event were displayed with a solid outline if results were out of range and without an outline if within the normal range.  A user could also click on an individual map point to view a time series of the 2021 data for the selected location that included upper and lower limits for the long-term monthly averages.  This functionality required additional modifications to the UI, in addition to developing custom code in the synthesis workflow that summarized the normal ranges prior to uploading data to the dashboard.  Ultimately, these changes improved the quality of information provided by the response-based monitoring by leveraging the decades of long-term monitoring data available for Tampa Bay.

@@

In addition to providing a centralized location for viewing results of the 2021 monitoring, the TBEP also had a need to understand how the broader community was engaging with the dashboard content. The Google Analytics service was used to track the number of users over time, their approximate location, how long a user engaged with the dashboard, and how the dashboard was accessed.  From May 13th when the analytics were added to September, `r totusr` users visited the dashboard, with about an even split of users accessing through a mobile device or a web browser on a PC.  Users were documented from `r demsum['City']` cities, `r demsum['State']` states, and `r demsum['Country']` countries (Table \@ref(tab:demtab)).  Understandably, users from the state of Florida accounted for a majority of site visits  An average of one minute and 21 seconds was spent on the dashboard by each user, although this time varied throughout the summer. The number of users also varied over time (Figure \@ref(fig:analytics)a), with the highest number of visits in July, which corresponded to severe red tide conditions, peak cell counts for *K. brevis*, and observable fish kills in densely populated areas of the bay [@Beck22].  Most users accessed the website through referrals from other websites (e.g., Figure \@ref(fig:analytics)b), although a substantial amount of visitors also accessed the dashboard directly (e.g., bookmarks or entering the URL in a web browser) or through email links.  Accessing the dashboard through social media or organically through a search engine were less common.  Trends in how users accessed the dashboard did not appear to change over time.

@@

The release of wastewater from Piney Point in the spring of 2021 mobilized dozens of local resource managers and stakeholders to track the potential impacts to Tampa Bay. The communication tools and workflows used to synthesize the data collected in this effort were useful for informing the management response and the analytics of our dashboard suggested that the products created by TBEP were valuable resources in 2021.  However, dashboard analytics are not the only metric of success and additional information can help assess the effectiveness of these tools to efficiently and transparently communicate science for decision-making.  Area researchers and managers were able to use screenshots from the dashboard to communicate results in real time, although there was no formal metric for these applications.  More holistically assessing the value of open science tools is also needed, such as how they may be used to inform policy or regulatory changes aimed at minimizing environmental impacts.

@@

The permanent closure of Piney Point in an environmentally responsible manner remains the ultimate goal of regional resource managers and policy-makers.  The open science tools created by the TBEP in 2021 were effective based on the limited means of assessing their impact, yet it is hoped that the tools can facilitate conversations towards broader goals.  In addition to Piney Point, mining activities in Florida and elsewhere continue to impose external costs on human health and the environment. The role of science in documenting these impacts is important, but even more so, research priorities should focus on understanding societal and political mechanisms that enable these preventable impacts to occur.  Combining traditional science with open science technologies can effectively enable positive change through better science communication.

@@

# Figures

```{r map, fig.cap = paste0('Response-based monitoring effort near Piney Point from March through September 2021 with (a) effort by monitoring group and (b) monitoring data type.  Monitoring groups included ', knitr::combine_words(grps), '. Monitoring data included algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality). Inset shows location of Tampa Bay on the Gulf coast of Florida, USA.')}
knitr::include_graphics(here('figs/map.jpeg'))
```

```{r synthesis, fig.cap = 'Data synthesis workflow for Piney Point data.  Raw data were uploaded by partners to Google Drive, data were synthesized using R-based tools, data and dashboard code were uploaded in a unified format on GitHub that included automated testing, and the app was deployed using Shiny Server.'}
knitr::include_graphics(here('figs/synthesis.JPG'))
```

```{r dashboard, fig.cap = 'Landing page for the Piney Point dashboard. Users can view a summary of effort to date and access additional tabs to view specific datasets.'}
knitr::include_graphics('figs/dashboard.JPG')
```

@@

# Tables

```{r efftab}
cap.val <- 'Sampling effort for response-based monitoring by partners from April 1st 2021 through September.  Values show the total number of unique sites sampled for water quality, algae (phytoplankton), and seagrass/macroalgae in Tampa Bay.  Values in parentheses show the percentage of the total for each monitoring type.'
knitr::kable(efftab, booktabs = T, caption = cap.val)
```

```{r demtab}
cap.val <- 'Top ten locations of users by city, state, and country for the Piney Point dashboard.  Numbers show the total users and percentages from the total for all locations.  Total cities, states, and countries that accessed the dashboard are shown in the column headers in parentheses.  Summaries are from May 13th, 2021 when analytics were included in the dashboard through September.'
knitr::kable(demtab, booktabs = T, caption = cap.val)
```

>>>>

The long-term recovery of Tampa Bay, Florida represents a collaborative, multi-stakeholder effort to quantify and limit external nitrogen loading as a key stressor impacting ecosystem health.  Nitrogen has been reduced by approximately two thirds of historical loading primarily through upgrades to wastewater treatment systems, stormwater controls, and habitat restoration projects [@Johansson91;@Lewis98;@Greening14;@Beck19].  As a result, chlorophyll concentrations have decreased, water clarity has improved, and seagrasses have recovered to a historical 1950s baseline by 2016 [@Greening14;@Sherwood17].  Despite these historical gains, ongoing management challenges for protecting and restoring the natural resources of Tampa Bay persist from continued threats of climate change and increasing watershed development.  Recent seagrass losses and seasonal harmful algal blooms highlight the potential impacts of persistent and increasing stressors affecting the bay [@Lopez16;@Tomasko20].

Impacts to human health and the environment associated with extractive mining activities are pervasive worldwide.  In Florida, phosphate mining has generated a multi-billion dollar fertilizer industry that supports agricultural production on the global market [@Henderson04]. Seventeen phosphate mining facilities occur in the Tampa Bay watershed and wastewater release into surface waters from these facilities has negatively impacted environmental resources.  The Piney Point facility on the southeast shore of Tampa Bay has been inactive for over twenty years and holding capacity to maintain wastewater onsite has decreased over time, contributing to releases in the early 2000s and in 2011 to Tampa Bay [@Garrett11;@Switzer11].  In March 2021, a leak was identified in the southern holding pond and wastewater release was authorized by the Florida Department of Environmental Protection (FDEP) to prevent catastrophic failure of the holding ponds.  Approximately 215 million gallons of mixed legacy phosphate mining wastewater and seawater from port dredging operations were discharged over a ten day period, introducing 205 tons of nitrogen to Tampa Bay [@Beck22].

@@

# Data collection

Monitoring of bay resources in response to release from Piney Point began in late March of 2021 and continued through the summer and early fall.  Several partners from public, private, and academic institutions conducted monitoring of water quality, phytoplankton (including the red tide organism *Karenia brevis*), seagrasses, macroalgae, and benthic habitats (Figure \@ref(fig:map)).  Monitoring partners included the FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute (FWRI) of the Florida Fish and Wildlife Conservation Commission (FWC), City of St. Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida (Figure \@ref(fig:map)a).  Coordination among monitoring agencies was facilitated by the TBEP under guidance of a plume simulation model run by the Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science [@Chen18;@Chen19]. Initial monitoring was focused near the release site at Port Manatee (Figure \@ref(fig:map)), but later expanded throughout middle and lower Tampa Bay, Boca Ciega Bay, Manatee River, and Sarasota Bay, as the simulation model predicted dispersal of the plume away from Piney Point [@Liu21].  Sampling effort was most intense in April, with less frequent sampling from May through September, although seagrass and macroalgae sampling was consistent throughout the event (Table \@ref(tab:efftab)).

Water quality monitoring focused on parameters that were expected to indicate a response to elevated nutrient loads from the Piney Point release.  Because Tampa Bay is a nitrogen-limited estuary [@Greening14;@Greening06;@Wang99], water quality monitoring focused primarily total nitrogen and inorganic forms (i.e., total ammonia nitrogen, nitrate/nitrite), chlorophyll-a as an indication of phytoplankton growth, and additional parameters expected to respond to nutrient loading (e.g., secchi depth, dissolved oxygen).  Phytoplankton monitoring included both qualitative (presence/absence) and quantitative (cells/L) summaries of taxa enumerated by microscopy from water samples taken at similar frequency and distribution as the water quality sites.  Phytoplankton data also included event-based sampling for *Karenia brevis*.  Finally, seagrass and macroalgae data were collected by transect sampling at locations covering the shoreline north of Piney Point and extending to the south into northern Sarasota Bay. Transect sampling included identification of seagrass species and macroalgae, typically to genus, and Braun-Blanquet abundance estimates at 10 m distances along each 50 m transect.  Details on all sampling methods are provided in @Beck22.

# Data synthesis

@@

Combining the source files on Google Drive into a synthesized format in R also required a consistent set of variables and naming convention that were applied to each file.  For example, parameters included in each water quality file were converted to a common set of names and all units were standardized (e.g., total nitrogen as `tn` and mg/L, chlorophyll-a as `chla` and $\mu$g/L, etc.). This required developing custom processing code for data files from each partner to address each unique use case.  As a result, the final processing file included almost 3000 lines of code to create the synthesized data products.  Each partner agency was also assigned a unique identifier in the synthesized files so effort could be tracked and reported in the dashboard (e.g., Pinellas County as `pinco`, Florida Department of Environmental Protection as `fldep`, etc.).  All synthesized files were saved in a long and tidy format, where each observation was given its own row, each variable was in its own column, and only one value was assigned to a cell [@Wickham14c].  This allowed for easily filtering observations of interest (as required with the dashboard) using the standardized naming conventions applied to each source file. More importantly, the developed workflow was used iteratively, often several times each week as new data files were uploaded to the Drive folders.  This required executing the synthesis code locally (i.e., on a personal computer) before pushing data to the dashboard and constantly revising the code to address new data formats or issues encountered when new files were uploaded.

The local workflow created by TBEP was hosted online in a GitHub repository [https://github.com/tbep-tech/piney-point, @Beck21d] that allowed use of additional resources that were critical to the synthesis effort (Figure \@ref(fig:synthesis), testing and sharing).  Hosting the RStudio project on GitHub served multiple purposes of 1) using version control to document and track changes using Git software, 2) making the code publicly accessible for transparency with the larger research and management community, 3) allowing automated testing of repository content, and 4) serving as an access point for deploying the dashboard on Shiny Server (see below).  The use of automated testing was an important service to identify potential errors or mistakes in the synthesized datasets. The testthat R package [@Wickham11] was used to write several checks for each dataset that were considered essential for maintaining integrity of the information.  Each time local changes were made to the workflow and pushed to the GitHub repository, tests were run automatically with GitHub Actions to verify each dataset included appropriate and accurate results.  For example, a test was used to verify that all water quality parameters were appropriately named and did not include any missing values. An automated notification was sent via email to TBEP staff if an error was encountered.  This service reduced the amount of time required to manually check the data and provided assurance that the reported results in the dashboard were correct.

@@

An important service provided by the dashboard was creating context for the response-based monitoring relative to baseline conditions in Tampa Bay.  This allowed dashboard users to quickly assess if the 2021 conditions were abnormal relative to typical seasonal patterns observed in the bay.  In addition to the baseline data tabs, all water quality data from the response-based monitoring were compared to the long-term monthly averages for data at the nearest ambient monitoring station [@Beck22].  Elements of the UI and summary plots allowed a user to quickly determine if the results were outside of the expected seasonal range.  Map points for a sampling event were displayed with a solid outline if results were out of range and without an outline if within the normal range.  A user could also click on an individual map point to view a time series of the 2021 data for the selected location that included upper and lower limits for the long-term monthly averages.  This functionality required additional modifications to the UI, in addition to developing custom code in the synthesis workflow that summarized the normal ranges prior to uploading data to the dashboard.  Ultimately, these changes improved the quality of information provided by the response-based monitoring by leveraging the decades of long-term monitoring data available for Tampa Bay.

@@

In addition to providing a centralized location for viewing results of the 2021 monitoring, the TBEP also had a need to understand how the broader community was engaging with the dashboard content. The Google Analytics service was used to track the number of users over time, their approximate location, how long a user engaged with the dashboard, and how the dashboard was accessed.  From May 13th when the analytics were added to September, `r totusr` users visited the dashboard, with about an even split of users accessing through a mobile device or a web browser on a personal computer.  Users were documented from `r demsum['City']` cities, `r demsum['State']` states, and `r demsum['Country']` countries (Table \@ref(tab:demtab)).  Understandably, users from the state of Florida accounted for a majority of site visits  An average of one minute and 21 seconds was spent on the dashboard by each user, although this time varied throughout the summer. The number of users also varied over time (Figure \@ref(fig:analytics)a), with the highest number of visits in July, which corresponded to severe red tide conditions, peak cell counts for *K. brevis*, and observable fish kills in densely populated areas of the bay [@Beck22].  Most users accessed the website through referrals from other websites (e.g., Figure \@ref(fig:analytics)b), although a substantial amount of visitors also accessed the dashboard directly (e.g., bookmarks or entering the URL in a web browser) or through email links.  Accessing the dashboard through social media or organically through a search engine were less common.  Trends in how users accessed the dashboard did not appear to change over time.

@@

The release of wastewater from Piney Point in the spring of 2021 mobilized dozens of local resource managers and stakeholders to track the potential impacts to Tampa Bay. The communication tools and workflows used to synthesize the data collected in this effort were useful for informing the management response and the analytics of our dashboard suggested that the products created by TBEP were valuable resources in 2021.  However, dashboard analytics are not the only metric of success and additional information can help assess the effectiveness of these tools to efficiently and transparently communicate science for decision-making.  Area researchers and managers were able to use screenshots from the dashboard to communicate results in real time, although there was no formal metric for these applications.  More holistically assessing the value of open science tools is needed, such as how they may be used to inform policy or regulatory changes aimed at minimizing environmental impacts.

@@

The permanent closure of Piney Point in an environmentally responsible manner remains the ultimate goal of regional resource managers and policy-makers.  The open science tools created by the TBEP in 2021 were effective based on the limited means of assessing their impact, yet it is hoped that the tools can facilitate conversations towards broader goals.  In addition to Piney Point, mining activities in Florida and elsewhere continue to impose external costs on human health and the environment. The role of science in documenting these impacts is important, but even more so, research priorities should focus on understanding societal and political mechanisms that enable these preventable impacts to occur.  Combining traditional science with open science technologies is one path to effectively enable positive change through better science communication.

@@

# Figures {-}

```{r map, fig.cap = paste0('Response-based monitoring effort near Piney Point from April through September 2021 with (a) effort by monitoring group and (b) monitoring data type.  Monitoring groups included ', knitr::combine_words(grps), '. Monitoring data included water quality (field-based and laboratory samples), algae sampling, seagrass and macroalgae, and mixed monitoring (water quality, algae, seagrass and macroalgae). Inset shows location of Tampa Bay on the Gulf coast of Florida, USA.')}
knitr::include_graphics(here('figs/map.jpeg'))
```

```{r synthesis, fig.cap = 'Data synthesis workflow for Piney Point data.  Raw data were uploaded by partners to Google Drive, data were synthesized using R-based tools, data and dashboard code were uploaded in a unified format on GitHub that included automated testing, and the dashboard was deployed using Shiny Server.'}
knitr::include_graphics(here('figs/synthesis.JPG'))
```

```{r dashboard, fig.cap = 'Landing page for the Piney Point dashboard. Users can view a summary of effort to date, when the data were last updated, and access additional tabs to view specific datasets.'}
knitr::include_graphics('figs/dashboard.JPG')
```

@@

# Tables {-}

```{r efftab}
cap.val <- 'Sampling effort for response-based monitoring by partners from April 1st 2021 through September.  Values show the total number of unique sites sampled for water quality, algae (phytoplankton), and seagrass/macroalgae in Tampa Bay.  Values in parentheses show the percentage of the total sites for each monitoring type across months.'
knitr::kable(efftab, booktabs = T, caption = cap.val)
```

```{r demtab}
cap.val <- 'Top ten locations of users by city, state, and country for the Piney Point dashboard.  Numbers show the total users and percentages from the total for each location type.  Total cities, states, and countries that accessed the dashboard are shown in the column headers in parentheses.  Summaries are from May 13th, 2021 when analytics were included in the dashboard through September.'
knitr::kable(demtab, booktabs = T, caption = cap.val)
```
